Interactive Model Serving for Regulatory Genomics

### 2016-12-16 paper.tex

% Copyright
\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

\title{Model Serving for Transcription Factor Binding Site Prediction}

@@

\maketitle
\begin{abstract}
Current pipelines for training machine learning models on genomic and epigenetic datasets lack resources to scale to current dataset sizes. Furthermore, pipelines for testing and evaluating these machine learning models are segmented, as they require users to separately test pipelines and evaluate their methods using disjunct software. In this paper we survey current workflows for training, testing and evaluating genomic machine learning models and provide an alternative solution to connecting the dots between these processes. More specifically, we present a complete fast and interactive pipeline that allows users to save prebuilt models and evaluate these models in a visual environment with low latency.
\end{abstract}

@@

\section{Introduction}
Regulatory genomics is a field of research that works to connect the function of DNA sequence, DNA interacting proteins and cell environment with gene expression. Understanding and predicting regulatory function is important, as it allows clinicians to predict the effect of an individual's genetic variants on their health and well being. Currently, there exist many different computational methods for prediction of regulatory function, including convolutional neural networks and SVM classification problems (Alipanahi et. al. 2014, Kelley et. al. 2016).  Datasets used to predict regulatory function are massive, reaching 15 TB in the ENCODE consortium (cite). However, these datasets are noisy, and thus analysis of regulatory predictions often require a human in the loop to visually interpret predictions against ground truth datasets produced in datasets such as ENCODE. \\

Current pipelines for training and analyzing regulatory genomic models include a segmented group of steps that are exclusive and require preprocessing between each stage. For example, the state-of-the-art pipeline involves executing a locally run algorithm, such as a kernel method or neural network, built for a single machine. Users who are interested in predicting on a particular sequence face multiple challenges. Because the state-of-the-art methods are not built to scale past a single machine, users must first sub-select a number of regions of the genome to train on specific to their hardware capabilities. After sub-selecting the data and training on that data, the user would then load these regions into a single node visualization tool, such as IGV, and repeat the preprocessing step to validate their model. A demonstration of the workflow pipeline required using currently available tools can be shown in Figure~\ref{fig:strawmanPipeline}.

@@

In the paper, we describe an end-to-end pipeline that efficiently computes and predicts on user provided sequences while maintaining seamless integration between processing steps. In our previous work, we use a scalable and parallel algorithm to efficiently compute protein binding sites without the use of specialized hardware (Morrow et. al. 2016). Here, we utilize this algorithm to predict from regulatory datasets, and provide a full analysis pipeline to visualize results. This novel pipeline allows users to interactively predict on regions of interest in a query genome. We compare our pipeline to the current state-of-the-art approaches on a single node, and then analyze pipeline latency at scale. \\

@@

\subsection{Algorithms}
Current algorithms for predicting on regulatory genomic datasets include neural networks and SVM classification. 
\begin{table}[t]
\centering
\caption{Featurization Techniques}
\label{my-label}
\begin{tabular}{| L{2cm} | L{1.45cm} | L{2cm} | L{1.75cm} |}
\hline
Featurization Technique & Datasets & Description & Usage
\\ \hline
Read Wavelets \newline 
(Raj et. al. 2015)           & DNA-seq, RNA-seq           & Iteratively quantifies read pileup                                 & Identification of genomic footprints \\ \hline 
Kernel Approximation  
(Morrow et. al. 2016) & Sequence, DNA-seq,  
RNA-seq & Approximates kernel matrix through random features & Generic, Widely applicable                                      \\ \hline
K-mer counts & Sequence & Counts the number of times each motif 
appears in a sequence of length k & Identification 
of motifs, quantification of sequence similarity \\ \hline
\end{tabular}
\end{table}
These methods train on small portions of the genome. For example, a neural net implementation DeepBind sub-selects 10,000 sequences to predict protein binding sites. Similarly, Kernel SVM uses 10,000 sequence to learn a classifier. Both tools are intended to run on a single machine. \\

\subsection{Evaluation Tools}
Main interactive tools for evaluating algorithms discussed above include visualization tools, such as University of Santa Cruz Genome Browser and Integrated Genomics Viewer (IGV). IGV is a desktop application that allows visualization of a wide variety of genomic datasets. IGV is used by clinicians and researchers to verify anomalies in their data and verify predictions to original raw genomic datasets. IGV does not support integrated visualization of predictive models, so users attempting to perform visual analysis of models using heterogeneous datasets must predict outside of the visualization tool.

@@

\subsection{ADAM}
ADAM is a framework for processing genomic data, and is built to run in the Spark ecosystem. ADAM uses distributed abstractions in Spark to batch process raw genomic datasets in parallel. More importantly, ADAM provides standardized schemas intended for various types of genomic datasets, presented in a narrow waist design that can be easily accessible by external applications. A figure of schema design is shown in Figure X. Important schemas that are used for epigenetic datasets include Alignment Record, Features, and Variant schemas. These schemas are demonstrated in Figure~\ref{fig:adam}. \\

@@

\subsection{Mango}
Mango is a distributed genome visualization browser that selectively materializes and organizes genomic data to provide fast in-memory queries.  Mango materializes data from persistent storage as the user requests different regions of the genome. This data is efficiently partitioned and organized in memory using a specialized RDD structure built upon interval trees. This interval based organizational structure supports ad hoc queries, filters, and joins across multiple files, enabling exploratory interaction with genomic data.  Mango is built on top of Spark and ADAM. A demonstration of Mango architecture can be seen in Figure X. Mango is implemented in three layers, namely, the cluster, server and client. In the cluster layer, Mango is built on ADAM and Spark, allowing users to perform and visualize fine grained queries on large genomic files. The server layer reduces and samples queries so that users can visualize their queries in an efficient and interpretable manner.

% \begin{figure}
%   \label{fig:mango}
%   \includegraphics[width=0.5\textwidth]{figures/mango.png}
%   \caption{Stack diagram for Mango architecture. Mango is built on ADAM and Spark, allowing users to perform and visualize fine grained queries on large genomic files.
%   Endive}
% \end{figure}

Our machine learning tool, called Endive, provides core abstract stages for building end-to-end machine learning pipelines to train and evaluate on genomic datasets. These stages include data preprocessing, featurization, and various solvers apt for genomic workloads. These stages in a machine learning pipeline can be mixed and matched to create novel pipelines tailored to a genomic prediction task of interest.
Preprocessing in Endive involves taking raw genomic data and parsing it into a vector of numeric values which can be directly read by a machine learning model. These preprocessing steps include formatting datasets as well as joining together heterogeneous datasets into one record based on coordinate location in the genome.
Featurization steps in Endive involve taking preprocessed datasets created in the preprocessing stage and transforming these records to explicit features that can be trained and tested on. Table 1 contains examples of these featurization steps. These computed features can then be fed into a choice of solvers developed in Keystone ML to train these machine learning models (Sparks, 2015).

\subsection{Machine Learning Pipeline for Epigenetic Workloads}
For the entirety of this paper, we will focus on using kernel approximations described in Table 1 to predict protein binding sites for an example pipeline.

\begin{figure}
  \label{fig:pipeline}
  \includegraphics[width=0.5\textwidth]{figures/systemDiagram}
  \caption{System Design overview.}
\end{figure}

\subsection{Interactive Prediction of Binding Sites}
Initially, direct prediction of binding sites was not in interactive time (6s). To reduce latency, we took advantage that regulatory regions are fairly sparse, occurring in less than 0.01\% of the entire genome. To take advantage of this sparsity, we iteratively analyzed the prediction of large potential regulatory regions with a pre-fetching technique. If this prediction scored higher than a given threshold, we cut the sequence into two equal parts iteratively pre-fetched the sequence cut in half, similar to a binary search implementation. This binary pre-fetching implementation  allowed us to significantly reduce average prediction time of a servable chunk. A schematic of this approach is seen in Figure~\ref{fig:iterations}.

\begin{figure}[b]
  \label{fig:iterations}
  \includegraphics[width=0.5\textwidth]{figures/iterations.png}
  \caption{Binary Pre-fetching}
\end{figure}

\section{Experiments}

For preliminary results, we train a model predicting transcription factor binding sites using the kernel approximation featurization method described in Table 1 from ENCODE datasets using DNA sequence.

\subsection{Fast Prediction with Mango}
For a straw man comparison, we took the state of the art pipeline used to train and analyze protein binding sites on a local computer. We trained a model on 15,000 DNA sequences to predict the location of binding sites of the EGR1 transcription factor, then visualizing the results in a genome browser, IGV. The state of the art in transcription factor binding is DeepBind, and uses a neural net. Previously, we demonstrated the improved accuracy and training time of the Convolutional Kitchen Sink over DeepBind's neural net implementation (Morrow et. al. 2016). The results of the runtime experiment for serving the model are displayed in Figure~\ref{fig:igvcomp}. A significant difference between Model Serving with Mango and using IGV is that with Mango, we are able to use Servable Chunks, which are defined in the user space. With IGV, there is no way to perform this prediction without completing the entire prediction over the whole search space.

@@

\subsection{Whole Genome Model Serving}
This model was solved using Keystone ML's Weighted Least Squares estimator, which accounts for large class imbalance of positive binding sites across the genome. This model was trained on 200,000,000 regions of the genome and generated a 4 MB model to be tested on new datasets. This model was loaded into the Mango genome browser and was used to interactively predict on DNA reference sequences as the user scrolled to new regions of the browser.

\begin{table}[b]
\centering
\caption{Predicting on Entire Genome}
\label{tableresults}
\begin{tabular}{|l|l|l|}
\hline
\textbf{Mean} & \textbf{Min} & \textbf{Max} \\ \hline
402.31ms      & 3.09ms       & 3.98s       \\ \hline
\end{tabular}
\end{table}

Table 2 shows the results of this experiment. Even on a significantly larger model, we averaged prediction times in the 400ms range. This is close to the level required for interactive applications, particularly applications with a user interface.

@@

\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case

\subsection{References}
\texttt{{\char'134}thebibliography}.

>>>>

\title{Interactive Model Serving for Regulatory Genomics}

@@

\maketitle
\begin{abstract}
Current pipelines for training machine learning models on genomic and epigenetic datasets lack resources to scale to current datasets sizes. Furthermore, pipelines for testing and evaluating these machine learning models are segmented, as they require users to separately test pipelines and evaluate their methods using disjunct software. In this paper we survey current workflows for training, testing and evaluating genomic machine learning models and provide an alternative solution to connecting the dots between these processes. More specifically, present a complete fast and interactive pipeline that allows users to save prebuilt models and evaluate these models in a visual environment with low latency.
\end{abstract}

@@

\section{Introduction}
Regulatory genomics is a field of research that works to connect the function of DNA sequence, DNA interacting proteins and cell environment with gene expression. Understanding and predicting regulatory function is important, as it allows clinicians to predict the effect of an individual's genetic variants on their health and well being. Currently, there exist many different computational methods for prediction of regulatory function, including convolutional neural networks and SVM classification problems \cite{alipanahi2015predicting}, \cite{kelley2016basset}.  Datasets used to predict regulatory function are massive, reaching 15 TB in the ENCODE consortium \cite{encode2004encode}. However, these datasets are noisy, and thus analysis of regulatory predictions often require a human in the loop to visually interpret predictions against ground truth datasets produced in datasets such as ENCODE. \\

Current pipelines for training and analyzing regulatory genomic models include a segmented group of steps that are exclusive and require preprocessing between each stage. For example, the state-of-the-art pipeline involves executing a locally run algorithm, such as a kernel method or neural network, built for a single machine. Users who are interested in predicting on a particular sequence face multiple challenges. Because the state-of-the-art methods are not built to scale past a single machine, users must first sub-select a number of regions of the genome to train on specific to their hardware capabilities. After sub-selecting the data and training on that data, the user would then load these regions into a single node visualization tool, such as \cite{igv2011}, and repeat the preprocessing step to validate their model. A demonstration of the workflow pipeline required using currently available tools can be shown in Figure~\ref{fig:strawmanPipeline}.

@@

In the paper, we describe an end-to-end pipeline that efficiently computes and predicts on user provided sequences while maintaining seamless integration between processing steps. In our previous work, we use a scalable and parallel algorithm to efficiently compute protein binding sites without the use of specialized hardware \cite{tfbinding}. Here, we utilize this algorithm to predict from regulatory datasets, and provide a full analysis pipeline to visualize results. This novel pipeline allows users to interactively predict on regions of interest in a query genome. We compare our pipeline to the current state-of-the-art approaches on a single node, and then analyze pipeline latency at scale. \\

\begin{figure}
  \label{fig:strawmanPipeline}
  \includegraphics[width=0.5\textwidth]{figures/strawman.png}
  \caption{Stawman pipeline for training and evaluating model.}
\end{figure}

Here, we provide an end-to-end pipeline that efficiently computes and predicts on regulatory genomics while maintaining seamless integration between processing steps. In our previous work, we use a scalable and parallel algorithm to efficiently compute protein binding sites on the whole genome, bypassing user defined subselection methods required for local algorithms \cite{tfbinding}. Here, we utilize this algorithm to interactively predict on new datasets, and provide a full analysis pipeline to analyze results. Our pipeline allows users to interactively predict on regions of the genome, removing file preprocessing steps required in current pipelines. We compare our pipeline to the current state of the art approaches intended for a local machine, and then analyze pipeline latency at scale on the whole genome.

@@

\subsection{Algorithms}
Current algorithms for predicting on regulatory genomic datasets include neural networks and SVM classification.
\begin{table}[t]
\centering
\caption{Featurization Techniques}
\label{table:featurization}
\begin{tabular}{| L{2cm} | L{1.45cm} | L{2cm} | L{1.75cm} |}
\hline
Featurization Technique & Datasets & Description & Usage
\\ \hline
Read Wavelets \newline
\cite{wavelets}         & DNA-seq, RNA-seq           & Iteratively quantifies read pileup                                 & Identification of genomic footprints \\ \hline
Kernel Approximation
\cite{tfbinding} & Sequence, DNA-seq,
RNA-seq & Approximates kernel matrix through random features & Generic, Widely applicable \cite{alipanahi2015predicting}                                       \\ \hline
K-mer counts & Sequence & Counts the number of times each motif
appears in a sequence of length k & Identification
of motifs, quantification of sequence similarity \\ \hline
\end{tabular}
\end{table}
These methods train on small portions of the genome. For example, a neural net implementation DeepBind sub-selects 10,000 sequences to predict protein binding sites. Similarly, Kernel SVM uses 10,000 sequence to learn a classifier. Both tools are intended to run on a single machine. \\

\subsection{Evaluation Tools}
Main interative tools for evaluating algorithms discussed above include visualization tools, such as University of Santa Cruz Genome Browser and IGV (\cite{ucscbrowser}, \cite{igv2011}). IGV is a desktop application that allows visualization of a wide variety of genomic datasets. IGV is used by clinicians and researchers to verify anomalies in their data and verify predictions to original raw genomic datasets. IGV does not support integrated visualization of predictive models, so users attempting to perform visual analysis of models using heterogeneous datasets must predict outside of the visualization tool.

@@

\subsection{ADAM: Distributed Genomic Workloads}
ADAM is a framework for processing genomic data, and is built to run in the Spark ecosystem \cite{adam}. ADAM uses distributed abstractions in Spark to batch process raw genomic datasets in parallel. More importantly, ADAM provides standardized schemas intended for various types of genomic datasets, presented in a narrow waist design that can be easily accessible by external applications. A figure of schema design is shown in Figure X. Important schemas that are used for epigenetic datasets include Alignment Record, Features, and Variant schemas. These schemas are demonstrated in Figure~\ref{fig:adam}. \\

@@

\subsection{Mango: Genome Visualization}
Mango is a distributed genome visualization browser that selectively materializes and organizes genomic data to provide fast and fine grained queries on datasets. Mango materializes data from persistent storage as the user requests different regions of the genome. This data is efficiently partitioned and organized in memory using a specialized RDD structure built upon interval trees. This interval based organizational structure supports ad hoc queries, enabling exploratory interaction with genomic data.  Mango is built on top of Spark and ADAM. A demonstration of Mango architecture can be seen in Figure X. Mango is implemented in three layers, namely, the cluster, server and client. In the cluster layer, Mango is built on ADAM and Spark, allowing users to perform and visualize fine grained queries on large genomic files. The server layer reduces and samples queries so that users can visualize their queries in an efficient and interpretable manner.

\begin{figure}
  \label{fig:mango}
  \includegraphics[width=0.5\textwidth]{figures/mango.png}
  \caption{Stack diagram for Mango architecture. Mango is built on ADAM and Spark, allowing users to perform and visualize fine grained queries on large genomic files.
  Endive}
\end{figure}

\subsection{Endive: Regulatory Machine Learning Abstractions}
Our machine learning tool, called Endive, provides core abstract stages for building end-to-end machine learning pipelines to train and evaluate on genomic datasets. These stages include data preprocessing, featurization, and various solvers apt for genomic workloads. These stages in a machine learning pipeline can be mixed and matched to create novel pipelines tailored to a genomic prediction task of interest.
Preprocessing in Endive involves taking raw genomic data and parsing it into a vector of numeric values which can be directly read by a machine learning model. These preprocessing steps include formatting datasets as well as joining together heterogeneous datasets into one record based on coordinate location in the genome.
Featurization steps in Endive involve taking preprocessed datasets created in the preprocessing stage and transforming these records to explicit features that can be trained and tested on. Table~\ref{table:featurization} contains examples of these featurization steps. These computed features can then be fed into a choice of solvers developed in Keystone ML to train these machine learning models \cite{keystone}.

\begin{figure}
  \label{fig:pipeline}
  \includegraphics[width=0.5\textwidth]{figures/ourpipeline.png}
  \caption{Pipeline overview. We eliminate preprocessing and file formatting steps.}
\end{figure}

\section{Experiments}
For preliminary results, we train a model predicting protein binding sites using the kernel approximation featurization method described in Table~\ref{table:featurization}  from ENCODE datasets using DNA sequence. This model was solved using Keystone ML’s Weighted Least Squares estimator, which accounts for large class imbalance of positive binding sites across the genome. This model was trained both locally on a small dataset as well as in a distributed environment on a large (200 million) dataset. These models generated a 4 MB model to be tested on new datasets. This model was loaded into the Mango genome browser and was used to interactively predict on DNA reference sequences as the user scrolled to new regions of the browser.

\subsection{Local Comparison}
For a straw man comparison, we took the state of the art pipeline used to train and analyze protein binding sites on a local computer. We trained a model on 15,000 DNA sequences to predict the location of binding sites of the EGR1 transcription factor, then visualizing the results in a genome browser, IGV. Because the models were too expensive to train locally, we train both straw man and our models on one machine with 24 Xeon processors, and 256 GB of RAM and 1 Nvidia Tesla K20c GPU. The state of the art in transcription factor binding is DeepBind, and uses a neural net. Previously, we demonstrated the improved accuracy and training time of the Convolutional Kitchen Sink over DeepBind's neural net implementation \cite{tfbinding}. The results of the runtime experiment for serving the model are displayed in Figure~\ref{fig:igvcomp}. A significant difference between Model Serving with Mango and using IGV is that with Mango, we are able to use Servable Chunks, which are defined in the user space. With IGV, there is no way to perform this prediction without completing the entire prediction over the whole search space.

@@

\subsection{Whole Genome Model Serving}
For our second experiment, we trained on the full genome, consisting of 200 million sequences consisting of 640 GB to predict binding sites for protein EGR1. We visualize all raw data related to the model prediction problem, specifically DNA sequence and Dnase hypersensitivity files. Training the model took 58 minutes parallized across 16 nodes with Intel E5-2670 2.6 GHz 8 core CPU, 256 GB RAM and 4 1-TB HDFS hard drives. The generated model was saved and loaded into Mango to visualize predictions on raw datasets. Minimum, maximum and mean times to predict on new sequences using the sparcity model explained above is show in Table~\ref{table:fullGenome}. As in the local experiments, we test on region sizes of 2000 bp, 2000 bp, 4000bp, 8000bp and 16000bp. \\

The mean exectution time of 402.31 ms meets interactive requirements of 500ms. However, maximum execution time of 3.98s is too high for interactive time.

\begin{table}
\centering
% \vspace{-0.4cm}
\caption{Minimum, maximum and mean prediction response times for transcription factor EGR1.}
\label{table:fullGenome}
% \hspace*{-0.96cm}
\begin{tabular}{|l|l|l|}
\hline
\textbf{Min} & \textbf{Max} & \textbf{Mean} \\ \hline
3.09 ms & 3.98 s & 402.31 ms  \\ \hline
\end{tabular}
\end{table}

Table~\ref{table:fullGenome} shows the results of this experiment. Even on a significantly larger model, we averaged prediction times in the 400ms range. This is close to the level required for interactive applications, particularly applications with a user interface.

@@

\bibliographystyle{abbrv}
\bibliography{paperRef}
